---
title: "Stat627-05"
author: "Zaiwei Liu"
date: "Dec 7, 2015"
output: html_document
---


## 1 Download Gapminder Data
```{r}
require(downloader, quietly = T)
downloader::download("https://raw.githubusercontent.com/jennybc/gapminder/master/inst/gapminder.tsv", "gapminder.tsv")
```


## 2 Write Analysed Data to File

In this part, we firstly drop the observations for oceania, then fit linear regressions on life expectancy for each country and calculate the MSEs. The outputs will be written to a .tsv file. 

```{r, warning=FALSE, message=FALSE}
# library the packages and read in the data
require(plyr, quietly=T)
require(dplyr, quietly=T)
require(boot, quietly=T)
require(ggplot2, quietly=T)
Data <- read.delim("gapminder.tsv")

# drop the "Oceania"
Data1 <- Data %>% tbl_df() 
Data2 <- Data1 %>% filter(!(continent %in% "Oceania")) %>% droplevels()
Reorder <- function(dat, fun) {
  dat_new <- dat
  dat_new$continent <- dat %>% with(reorder(continent, gdpPercap, fun))
  dat_new
}
Data3 <- Reorder(Data2, "median")


# fit linear regressions on life expectancy and calculate the MSE for each country
LOOCV <- function(dat, offset=1952) {
  Model <- glm(lifeExp ~ I(year-1952), family=gaussian, dat)
  MSE <- cv.glm(dat, Model)$delta[1]
}
MSE <- ddply(Data3, ~ country, LOOCV)
names(MSE)[2] <- "mse"
Data4 <- left_join(MSE, Data3, by="country")
Data4 <- tbl_df(Data4)
write.table(Data4, "gapminder_new.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
```


## 3 Save Two Figures to one PDF

It seems that we're required in this assignment to save the figures into "one file". Since `ggsave` cannot deal with multiple plots, we decide to make use of the `multiplot` function from [Cookbook for R](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/). In the output, two plots will be merged into one single PDF.

Note that before making plots, the level of the factor variable `continent` is reordered according to the calculated MSE. 

```{r, warning=FALSE, message=FALSE}
require(plyr, quietly=T)
require(dplyr, quietly=T)
require(ggplot2, quietly=T)
Dat <- read.delim("gapminder_new.tsv")

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

# reorder the factor "continent" based on the MSE in each continent
arrange_order <- function(dat, type) {
  dat1 <- dat
  if (type==1) dat1 <- dat %>% arrange(mse)
  else if (type==2) dat1 <- dat %>% mutate(continent = reorder(continent, mse, median))
  else if (type==3) dat1 <- dat %>% mutate(continent = reorder(continent, mse, median)) %>% arrange(continent)
  else dat1 <- dat %>% mutate(continent = reorder(continent, mse, median)) %>% arrange(mse)
  dat1
}
Dat <- arrange_order(Dat, type=4)

Plot_a <- ggplot(Dat) + geom_violin(aes(continent,log(gdpPercap),color=continent)) +
  xlab("Continent") + ylab("Log-transformed GDP Per Capita") + 
  ggtitle("Log-transformed GDP Per Capita in Different Continents") + 
  stat_summary(aes(continent, log(gdpPercap)), fun.y="median", geom="point")
Plot_b <- Dat[,c(1,2,3)] %>% unique() %>% ggplot(aes(x = continent, y = log(mse), group = 1)) + 
  geom_jitter(aes(color=continent), position = position_jitter(width = .1)) + 
  stat_summary(fun.y="median", geom="line") + 
  ggtitle("Plot for Type I")
pdf("gapminder_ggplots.pdf")
multiplot(Plot_a, Plot_b)
graphics.off()
```