---
title: "Stat627-04"
author: "Zaiwei Liu"
date: "Nov 30, 2015"
output: html_document
---

## 0. Loading dirty Gapminder

First, let's read in the data with `strip.white = FALSE` and `strip.white = TRUE` and take a look at their differences. 

```{r Read in Data and Packages}
suppressPackageStartupMessages({
  library(plyr)
  library(dplyr)
})
Dirty1 <- read.delim('https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear_dirty.txt', strip.white = FALSE)
Dirty2 <- read.delim('https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear_dirty.txt', strip.white = TRUE)
str(Dirty1)
str(Dirty2)
```

With `strip.white = TRUE`, the leading and trailing white space from unquoted character fields in the dataset will be striped. If we set the argument to be `FALSE`, we can see that for the factor variable `region`, totally `r length(levels(Dirty1$region))` levels were read in, including several unexpected levels like "". Now if we set the argument to be `TRUE`, then the number of levels is reduced to `r length(levels(Dirty2$region))`, which is easier for us to deal with.

## 1. Splitting and Merging

Let's take a look at the columns of the dataframe. 

```{r}
head(Dirty2)
```

We can find that the `continent` and `country` variables are now pasted in one column, and we need to split it into two.

```{r Splitting}
### Split the factor column into two character columns and return as a new data.frame
SPLIT <- function (x) {
  # split
  tmp <- x %>% as.character() %>% strsplit(split='_') %>% 
    unlist() %>% matrix(ncol = 2, byrow = T) %>% data.frame()
  # assign the colnames
  colnames(tmp) <- c('continent','country')
  # transform the factor into char
  tmp$continent <- tmp$continent %>% as.character()
  tmp$country <- tmp$country %>% as.character()
  return(tmp)
}

Dirty <- SPLIT(Dirty2$region)
Dirty <- cbind(country = Dirty[,2], Dirty2[1:2], continent = Dirty[,1], Dirty2[,3:4])
head(Dirty)
```

Now, each of the columns only contains one conceptual variable.

## 2. Missing Values

Assuming that the missing values here are coded as either `""` or `NA`, a function is written to check if there's any missing value in a character vector. The return value here would be the indices of the missing values.

```{r Missing values}
### Check in a character vector if there's missing value in it, and return the index
Index_Missing <- function(x) {
  Index <- NULL
  for (i in 1:length(x)) {
    # we assume the missing values here are coded as either "" or NA
    if (x[i] == '' | is.na(x[i])) Index <- c(Index,i)
  }
  return(Index)
}

(Index <- union(Index_Missing(Dirty$continent), Index_Missing(Dirty$country)))
Dirty[Index,]
```

It's identified that rows `r Index` have missing fields in it. Now we're going to write a function to fill them with correct values. The idea is to replace the missing value with the most frequent alternative value for the same country. 

```{r}
### Fill the missing continent with the information of the other rows  
Fill_Continent <- function(x = Index) {
  Dat <- Dirty
  # get the name of all the countries with missing continent 
  Names <- unique(Dat[x, ]$country)
  for (i in 1:length(Names)) {
    # for a certain country, find the rows with the same country name
    tmp <- grep(pattern = Names[i], x = Dat$country, fixed = TRUE)
    # get the row number of the missing value (for this country)
    k <- intersect(Index, which(Dat$country == Names[i]))
    # replace the missing continent with the most frequent alternative
    Dat$continent[k] <- Dat$continent[tmp] %>% table() %>% which.max() %>% names()
  }
  return(Dat)
}

Dirty <- Fill_Continent(Index)
Dirty[Index,]
```

We can see that now the three rows are filled with the correct value. 

## 3. Inconsistent Capitalization and Spelling
Here, we may assume that the correct coding is highly possible to be the most frequently occured one, since for any certain kind of irregular spelling, inconsistent capitalization or typo, the chance of it happening very often is quite low in the sense of law of large number. 

Thus in this case, we just replace all of the irregular spellings with the most frequent spelling for each country. Of course, this rule of thumb might be violated sometimes, but here it works well for the gapminder data.

Note that even if this procedure fails, it's still a good point to get started and will certainly reduce some of the unnecessary use of regular expressions. 

As for the details of the coding, an easy for loop is used here, for the fact that the rows are well-ordered so that every 12 consecutive rows are of one country, and the countries are in alphabet order. 

```{r spelling}
tmp <- NULL
for (i in 1 : (dim(Dirty)[1]/12)) {
  tmp <- c(tmp, Dirty$country[(i*12-11):(i*12)] %>% table() %>% 
             which.max() %>% names() %>% rep(12))
}
Dirty$country <- as.factor(tmp)
Dirty <- Dirty %>% droplevels()
```

## 4. Final Check

The check shows that now the data is identical with the clean one. 

```{r}
Clean <- read.delim('https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear.txt')
identical(Dirty, Clean)
```

After these operations, the dataset is now successfully cleaned, and we didn't make too much use of regular expressions here. To show that I'm not afraid of using RegEx (I'm not sure if that counts for part of the grade...), a little example would be given as the end of this assignment. 

```{r}
eg <- SPLIT(Dirty2$region)
country <- gsub("\\b([a-z])", "\\U\\1", eg$country, perl = TRUE)
country <- gsub("\\b(And)", "\\L\\1", country, perl = TRUE)
country <- gsub("\\b(Of)", "\\L\\1", country, perl = TRUE)
country <- gsub("\\b(The)", "\\L\\1", country, perl = TRUE)
country <- gsub("\\b(D'Ivoire)", "d'Ivoire", country, perl = TRUE)
country <- gsub("\\b(D'Ivore)", "d'Ivoire", country, perl = TRUE)
country <- gsub("Democratic Republic of the Congo", "Congo, Dem. Rep.", country, perl = TRUE)
country <- gsub("Congo, Democratic Republic", "Congo, Dem. Rep.", country, perl = TRUE)
```

Now the column of `countries` becomes clean again, but look, isn't it a little bit weird and unnatural? 