---
title: "Stat627-03"
author: "Zaiwei Liu"
date: "Nov 1, 2015"
output: html_document
---
  
# 0 Prepare the Data and Packages  
  
```{r Read in Data, warning=FALSE, message=FALSE}
Data <- read.delim(file = "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt")
require(plyr, quietly=T)
require(dplyr, quietly=T)
require(ggplot2, quietly=T)
options(scipen=999)  # Disable scientific notation (the default value is scipen=0)
```

```{r Multiple Plot Function, echo=FALSE, results="hide"}
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# 1 Factor Management

## (a) Drop Oceania.
For the fact that there're only two countries in Oceania, we'd like to drop it from our analysis here. (Sorry for Australia and New Zealand but this is what is required in this part.)

Concrete information on the data structure will be provided to learn the effect of the functions.

### i. Before removing the rows

```{r Droplevels1}
Data <- Data %>% tbl_df()  # converts data to tbl class
str(Data)
```

We can see that before removing the rows, there're totally `r nrow(Data)` rows of observation in the data set, while the factor `country` has `r length(levels(Data$country))` levels and `continent` has `r length(levels(Data$continent))` levels. 

### ii. After removing the rows

```{r Droplevels2}
Data1 <- Data %>% filter(!(continent %in% "Oceania"))
str(Data1)
```

After removing the rows from Oceania, the number of rows now becomes `r nrow(Data1)`, while the levels of the two factors are still `r length(levels(Data1$country))` and `r length(levels(Data1$continent))` separately.  

### iii. After Dropping Levels

```{r Droplevels 3}
Data2 <- Data1 %>% droplevels()
str(Data2)
```

After applying `droplevels()` function to the data, then the numbers of levels of the affected factors change. Now, the level of `country` turns to be `r length(levels(Data2$country))` and the level of `continent` turns to be `r length(levels(Data2$continent))`.

### iv. Figures

Now, we're going to use figures with legends to show the effects of filtering and dropping levels of the data. The first one is the original violin plot we did last time, without any reordering, transformation and summaries. The second one, as a comparison, is a rather well-designed violin plot, with Oceania dropped, log-transformed, continent reordered, and the median GDP per capita is also plotted as a point in each case.  

```{r Plots, fig.height=5, fig.width=14}
Reorder <- function(dat, fun) {
  dat_new <- dat
  dat_new$continent <- dat %>% with(reorder(continent, gdpPercap, fun))
  dat_new
}
Data3 <- Reorder(Data2, "median")

A <- ggplot(Data) + geom_violin(aes(continent, gdpPercap, color=continent)) +
  xlab("Continent") + ylab("GDP Per Capita") + 
  ggtitle("GDP Per Capita in Different Continents")
B <- ggplot(Data3) + geom_violin(aes(continent,log(gdpPercap),color=continent)) +
  xlab("Continent") + ylab("Log-transformed GDP Per Capita") + 
  ggtitle("Log-transformed GDP Per Capita in Different Continents") + 
  stat_summary(aes(continent, log(gdpPercap)), fun.y="median", geom="point")
multiplot(A, B, cols=2)
```

Apparently, the second plot is more natural and informative compared to the first one.

## (b) Reorder Factors 

Last time, we've fitted linear regressions and got the LOOCV MSE for each country. For this part, we're going to reorder the factor `continent`, based on the median MSE in each continent. As suggested on JB's website, there're four ways to arrange/reorder the factor, and we're going to integrate them into a single principled funciton.  

```{r MSE Calulation Function, echo=FALSE, results="hide"}
require(boot, quietly=T)
LOOCV <- function(dat, offset=1952) {
  Model <- glm(lifeExp ~ I(year-1952), family=gaussian, dat)
  MSE <- cv.glm(dat, Model)$delta[1]
}
MSE <- ddply(Data, ~ country, LOOCV)
names(MSE)[2] <- "mse"
MSE <- left_join(MSE, unique(Data[,c(1,4)]), by="country")
MSE <- tbl_df(MSE)
```

```{r Arrange and Reorder}
arrange_order <- function(dat, type) {
  dat1 <- dat
  if (type==1) dat1 <- dat %>% arrange(mse)
  else if (type==2) dat1 <- dat %>% mutate(continent = reorder(continent, mse, median))
  else if (type==3) dat1 <- dat %>% mutate(continent = reorder(continent, mse, median)) %>% arrange(continent)
  else dat1 <- dat %>% mutate(continent = reorder(continent, mse, median)) %>% arrange(mse)
  dat1
}
MSE1 <- arrange_order(MSE, type=1)
MSE2 <- arrange_order(MSE, type=2)
MSE3 <- arrange_order(MSE, type=3)
MSE4 <- arrange_order(MSE, type=4)
```

### i. Post_arrange

The first type, simply arranging the data, will just mutate the order of the rows, but won't change the order of the factor `continent`. Such an arrangement will have no effect on the figure. 

```{r Demonstration1, results="hold"}
MSE1 %>% head()
MSE1 %>% str()
```

### ii. Post_reorder

The second type, simply reordering the factor, will not change the order of the rows, but the order of the factor `continent` will be changed. Such a reordering will change the plot.

```{r Demonstration2, results="hold"}
MSE2 %>% head()
MSE2 %>% str()
```

### iii. Post_both

The third type firstly reorders the factor then arranges the data based on the order of the factor. In this way, the rows of data are organized in chunks of continents, where countries in `Oceania` are on placed on the top since they have the least median MSE. The plot will be the same as the second type.  

```{r Demonstration3, results="hold"}
MSE3 %>% head()
MSE3 %>% str()
```

### iv. Swap out the argument for `arrange()`

The last type also reorders the factor first, but then arranges the data based on the mse of each country. In this way, the data are organized in ascending order of MSE. Personally, this is the one I preferred to use, and the plot here will also be the same as the second and third type.  

```{r Demonstration4, results="hold"}
MSE4 %>% head()
MSE4 %>% str()
```

### v. Figures



```{r figures, fig.height=5, fig.width=14}
A <- MSE1 %>% ggplot() + 
  geom_jitter(aes(x = continent, y = log(mse), color=continent), position = position_jitter(width = .1)) + 
  stat_summary(aes(continent, log(mse)), fun.y="median", geom="point") + 
  ggtitle("Plot for Type I")
B <- MSE4 %>% ggplot() + 
  geom_jitter(aes(x = continent, y = log(mse),  color=continent), position = position_jitter(width = .1)) + 
  stat_summary(aes(continent, log(mse)), fun.y="median", geom="point") +
  ggtitle("Plot for Type I, II & III")
multiplot(A, B, cols=2)
```

# 2 Visualization design

In this part, we remake the MSE plot we did in the last assignment, with a brandnew design of visualization. Now, the data are log-transformed, reordered, and represented in facets. The function here is a generalization of the situation, by passing the arguments to `ggplot`.   

```{r Reorder Country, fig.height=15, fig.width=12}
MSE_plot <- function (dat, ord_fac, quant, group) {
  X <- colnames(dat)
  X[which(X==ord_fac)] <- "ord_fac"
  X[which(X==quant)] <- "quant"
  X[which(X==group)] <- "group"
  colnames(dat) <- X
  PLOT <- ggplot(dat) + facet_grid(.~group) + 
    geom_point(aes(x = log(quant), y = reorder(ord_fac, log(quant)), color=group))
  PLOT
}
MSE_plot(MSE4, ord_fac = "country", quant = "mse", group = "continent") + 
  xlab("Log-transformed MSE") + ylab("Countries") +
  ggtitle("Log-transformed MSE for Countries") + theme(legend.position="none")
```


